Circularity in Voxel-Based Analysis

### 2011-10-19 ANTsTBSS/paperRevision.tex
ENH:  Removed unnecessary phrase.
Take a break, Driver 8

\begin{abstract}
Recent discussions within the neuroimaging community have highlighted the
problematic presence of selection bias in experimental design.  Although 
initially centering around the selection of voxels during the course of
fMRI studies, we demonstrate how this concept of circularity bias can 
potentially corrupt voxel-based analyses with a particular emphasis on
population studies of fractional anisotropy (FA).  
For such studies, template-based registration plays a critical role
in which a representative template serves as the normalized space for 
group alignment.  Following 
computation of FA, each subject's FA map is registered to the FA template prior to 
performing statistical comparisons between different groups within a population.
We illustrate how this analysis strategy can induce selection
bias which varies in strength with the similarity metric used in
registration.  Specifically, we show that the popular sum of squared
difference intensity metric  maximizes effect size, rather than
anatomical alignment, and consistently overestimates statistical
significance.  Consequently, we advocate a more cautious approach
that normalizes a population of FA images through the associated 
anatomical images.  One possibility, the Symmetric Group Normalization
strategy for unbiased template construction from anatomical (e.g.
T1-weighted) images, provides a valid and unbiased coordinate system for
statistical inference of FA differences.  
We establish this
outcome by using simulated data derived from normal controls from the International
Neuroimaging Data-sharing Initiative as well as a cohort of 16 traumatic
brain injury survivors compared with 17 control subjects.  
Alignment
comparisons with the normalization component of the popular tract-based
spatial statistics (TBSS) show that our anatomically-based template
approach produces good alignments despite the absence of any direct
FA-to-FA registration.  
%Of practical consideration we note that the
%software tools discussed in this work are provided to the public as part
%of the open source Advanced Normalization Tools (ANTs) repository.  
In conclusion, we recommend against circular analysis strategies that 
select image similarity metrics that are tied directly to the statistical 
functions used to assess population effects. 
\end{abstract}

@@

However, suppose the SSD metric is used to find a set of $M$ transformations that map a population of $M$ FA images, $\{I_1, I_2, \ldots, I_M\}$, to a representative FA template, $J$, for statistical testing between groups. This results in the transformations,
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations (since the transformations are mutually independent) and recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the mean of the aligned images, it becomes apparent that the transformation solution
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}}
\end{align}
is that which minimizes the voxelwise group variance, or at least an approximation thereof.  Reducing the group variance will also directly affect the value of a standard population statistic, the Student's $t$-test.  Other standard similarity metrics will also exhibit this
trend to some degree as they minimize some local measure of intensity-based distance between the subject and the template.

@@

Although various processing choices mitigate the effects of such bias,
such as Gaussian smoothing, nonparametric testing \citep{Rorden2007}, 
statistical analysis based on orthogonal projections onto the white matter skeleton 
\cite{Smith2006}, multiple comparisons corrections \cite{Nichols2003}, and the use of more robust similarity metrics, 
such choices are often made ad hoc and/or post hoc and, to our 
knowledge, not with the realization of the potential
for circularity bias described by Eqn. (\ref{eq:variance}).

Instead of direct FA-to-FA template mapping for 
alignment to the normalized space, we advocate a statistically 
independent alignment protocol in which the FA images are mapped to 
the common reference space via intra-subject FA-to-T1 mapping 
followed by T1-to-T1 template alignment thus eliminating all direct
FA-to-FA registrations. The key point is that the intensity value differences used 
to assess alignment (i.e. the similarity metric) in FA-to-T1 and T1-to-T1 
normalizations are statistically independent of the intensity value differences used for 
the post-alignment numerical analysis.  It should be noted that this discussion
is certainly not limited to studies involving FA.  From 
Eqn. (\ref{eq:variance}), it is evident that such bias can occur anytime the 
similarity metric (and the values employed) are non-independent of 
the statistical analysis.

Using simulated data, we first establish that this effect is real 
and warrants concern on the part of researchers using 
these techniques.  We then show how the general application
of a high performance \citep{Klein2009}, nonrigid registration algorithm 
\citep{Avants2011}, which forms the underlying machinery for an optimal 
template construction algorithm \citep{Avants2010}, can facilitate alignment 
between DTI-derived scalar images for cross-population studies.  
Although this type of alignment procedure is not uncommon 
(e.g. \cite{Lipton2008}), to our knowledge, this is the 
first time a connection has been made with the concept of 
circularity.

%The rest of the paper establishes the severity of 
%this bias in situations
% although scientific best 
%practices would dictate the reasonable removal of all 
%potential sources of bias (including circularity) regardless of 
%the general degree of severity.

\section{Methods and Materials}
% Provide sufficient detail to allow the work to be reproduced. 
% Methods already published should be indicated by a reference: 
% only relevant modifications should be described.

\subsection{Template Construction for Anatomically-Based Population Normalization}

\begin{figure*}
\begin{center}
\begin{tabular}{c}
  \includegraphics[width=170mm]{workflow.pdf}
%  \includegraphics[width=130mm]{workflow.pdf}
\end{tabular}
\caption{Possible ANTs-based workflow using SyGN for normalization of images for statistical analysis.  In contrast to standard  FA-to-FA template registration
(transform represented with the blue arrow), our proposed modification first constructs an ANTs optimal anatomical template as illustrated in the center of the figure.  Each FA image is then rigidly mapped, with distortion correction, to the individual T1 using the white matter mask as shown on the left.  The individual T1-weighted image is then nonrigidly registered to the anatomical template.  Optionally, one can register the T1 template to the MNI152 template which resides in the same space as the FMRIB58 template.  Thus, the composite transform, carefully constructed without any FA-to-FA image registration, can take an individual subject's FA map (or other DTI-derived measure) to the desired standardized coordinate system.    }
\label{fig:workflow}
\end{center}
\end{figure*}

In general, normalization to a standard coordinate system to
reduces intersubject variability in population studies.  Various
approaches exist for determining the normalized space such as selection
of a pre-existing template based on a single subject, e.g. the Talairach
atlas \citep{Talairach1988}, or a publicly available averaged group of
subjects, e.g. the MNI \citep{Collins1994} or ICBM \citep{Mazziotta1995}
templates.  
%One challenge with standard templates is that they may inadvertantly bias one's results by enabling better normalization of subjects to which the template is more similar.  This issue is exacerbated when dealing with populations that have high variance (e.g. due to disease) and/or when one's normalization method is low-dimensional (not flexible enough to capture large shape differences). 
Population-specific templates (or optimal templates) alleviate some of
these issues by deriving a most representative image from the population
\citep{Good2001}.  Large deformation registration algorithms also reduce
this confound by being less sensitive to the deformation distance
between subject and target.  Some approaches combine both advantages,
for instance, the diffeomorphic approach of Joshi et al. employs the
SSD metric and a shape distance to bring the subject group of images
into alignment \citep{Joshi2004}.  Variants
include extension to multiple modalities \citep{Lorenzen2006} and small deformations
\citep{Geng2009}.  These approaches iteratively minimize group difference in ``congealing''
towards a representative image template \citep{Learned-Miller2006}.
In contrast, Avants et al. explicitly model the geometric component of the 
normalized space during optimization.  Coupling the intrinsic symmetry of 
SyN pairwise registration \citep{Avants2011} and an
optimized sharpening/averaging of the template appearance, Symmetric Group
Normalization (SyGN) is a powerful framework for producing optimal population-specific
templates \citep{Avants2010} with arbitrary similarity metric choice.

%Given a set of representative images, 
%$\{I_1, I_2, \ldots, I_M\}$, optimization involves finding the set of paired
%diffeomorphic transformations, $\left\{\left(\phi^1_1, \phi^1_2\right), 
%\left(\phi^2_1, \phi^2_2\right), \ldots, \left(\phi^M_1, \phi^M_2\right) \right\}$,
%the optimal template appearance, $J^*$, and corresponding coordinate system, $\psi(\mathbf{x})$,
%which minimize the following cost function:
%\begin{align}
%  \sum_{m=1}^M \left[
%    D\left( \psi(\mathbf{x}), \phi^m_1(\mathbf{x}, 1) \right) + 
%    \Pi \left( I_m\left(\phi^m_2(\mathbf{x}, 0.5) \right), J^*\left(\phi^m_1(\mathbf{x}, 0.5) \right) \right)
%    \right]
%\end{align}
%where $D$ is the diffeomorphic shape distance, 
%\begin{align}
%  D\left(\phi(\mathbf{x}, 0), \phi(\mathbf{x}, 1)\right) = \int_{0}^1 \| v(\mathbf{x}, t) \|_L dt, 
%\end{align}
%dependent upon the choice of the linear operator, $L$, and
%$v$ is the diffeomorphism-generating velocity field, 
%\begin{align}
%  v\left(\phi(\mathbf{x}, t), t \right) = \frac{d\phi(\mathbf{x}, t)}{dt},\,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
%\end{align}
%$\Pi$ is the choice of similarity metric, often cross-correlation \citep{Avants2008a}, calculated in the 
%virtual domain midway between each individual image and the current estimate of the template. 
%
%With initial assignment of $\left\{\left(\phi^m_1, \phi^m_2\right)\right\}$ and $\psi(\mathbf{x})$ 
%to identity, iterative optimization
%involves estimating the pairwise transformations, estimation of the optimal template appearance, and 
%updating $\psi(\mathbf{x})$ by averaging the current estimate of $\left\{\phi_1^m\right\}$.  Additional details are given in \cite{Avants2010}.  
%For the interested reader, a set of scripts has been created and is distributed with the ANTs library for facilitating template creation.

Our proposed workflow applied to anatomically-based population normalization is illustrated in Fig. \ref{fig:workflow}.  Instead of direct FA-to-FA registration, our ANTs-based modification can be described as follows:

\begin{enumerate}
  \item Build the anatomical template from  the population sample or sample subset using SyGN as described in
  the previous section.
  \item If only using a sample subset for template creation, register the remaining
        anatomical data to the template and save the resulting transformation for each         
        subject indexed by $i$ (i.e. $\mathcal{T}^i_{template}$).
  \item Find the optimal transformation from each
        subject's FA to the anatomical template:
  \begin{enumerate}
  \item For each anatomical image, segment the white matter (in ANTs, one can use N4 bias correction \citep{Tustison2010} and Atropos $n$-tissue segmentation \citep{Avants2011a}).
  \item Rigidly transform each subject's average DWI to the bias-corrected anatomical image ($\mathcal{R}^i_{dwi}$).
  \item Perform distortion correction on the FA to anatomical mapping 
        by localized nonrigid registration
        of rigidly transformed FA map ($=\mathcal{R}^i_{dwi} \mathrm{FA}$) using
        the white matter mask obtained in step 3(a) ($\mathcal{T}^i_{FA}$).
  \item For each subject, concatenate the transforms which maps each
        subject's FA to the template ($\mathcal{T}^i_{total} = \mathcal{T}^i_{template} \circ 
        \mathcal{T}^i_{FA} \circ \mathcal{R}^i_{dwi} \mathrm{FA}$).
  \item Warp each subject's FA map to the template using $\mathcal{T}^i_{total}$.  
  Equivalently, one could warp the mean diffusion (MD), axial diffusion (AD) or any 
  other scalar derivable from DTI to the template.  
%  One other detail is that we prefer to warp the subject's DTI to 
%  the template and then, in template space, derive the scalar images of interest.  
  \end{enumerate}      
%  \item Perform the remaining portions of the standard TBSS pipeline:
%    \begin{enumerate}
%      \item Create the mean FA image from all template-aligned FA images.
%      \item Produce the white matter skeleton from the mean FA image.
%      \item Create skeleton distance map and project all FA data onto the skeleton.
%      \item Run the statistical analysis.
%    \end{enumerate}
\end{enumerate}
%The next section compares this framework to FA-based normalization strategies with respect to both a patient-control study and a study of gender effects in a control population.

@@

\subsubsection{NKI/Rockland Data for Generation of the Simulated DTI Gender Cohort}
Data from the first 14 weeks of prospective data-sharing initiative sponsored by the 1000 Functional Connectomes Project (FCP)%
\footnote{
In support of open science, the 1000 Functional Connectomes Project (FCP---http://fcon\_1000.projects.nitrc.org) 
was initiated on December 11, 2009 by various members of the MRI community \citep{Biswal2010}.  Motivated by the absence of neuroimaging data for research, this initiative seeks to form collaborative partnerships with imaging institutions for sharing well-documented multi-modal image sets accompanied by phenotypic data.  Commitments from institutions such as the NYU Institute for Pediatric Neuroscience and the Nathan Kline Institute (NKI)--Rockland have resulted in prospective data repositories
currently distributed to the public via the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC---http://www.nitrc.org).
}
were downloaded on January 15, 2011 comprised of 74 subjects (average age in years: 32.4 $\pm$ 17.8).  
Due to various issues (e.g. lack of corresponding T1 image, failed DTI reconstruction, and age matching requirements) only 62 subjects (21 females and 41 males).  Each imaging session for each subject produced a 64-directional diffusion tensor imaging scan (parameters: conventional single-shot spin echo
EPI pulse sequence, TR = 10000 ms, TE = 91 ms, axial acquisition, 
voxel size = $2 \times 2 \times 2$ mm$^3$) and a T1 anatomical scan (parameters:  TR = 2500 ms, b = 1000 s/mm$^2$ for each direction, 58 contiguous slices of 2.0 mm thickness,
TI = 1200 ms, TE = 3.5 ms, flip angle = 8$^\circ$, 192 contiguous slices of 
1.0 mm thickness, FOV = 256 $\times$ 256 mm$^2$, and voxel size = 1 mm$^3$).  Images were anonymized including defacing of the T1 images prior to upload. Diffusion tensor data associated with each of the following data sets were reconstructed from the diffusion weighted sequences using Camino%
\footnote{
http://camino.org.uk/
}---an open source toolkit for diffusion MRI processing and 
analysis \citep{Cook2006} in combination with ANTs-based 
registration for motion correction of the DTI sequence.

@@

\subsection{Simulated DTI Gender Cohort Creation}

@@

To provide realistic ground truth FA data, we implemented 
the work of Van Hecke et al. \cite{van-hecke2009} which 
we have also made available to the public
\cite{Tustison2011}.%
\footnote{
http://hdl.handle.net/10380/3315
}  
Given an input DTI atlas, b0 image, scheme file indicating
the gradient directions, and 
a cohort of aligned DTI data (to model intersubject variability), the
user is able to create 
simulated data sets with specified pathology and Rician noise.
For the simulated experiments described in the next section we
created a T1 template from 30 randomly chosen T1 images.  The
template was then registered to the MNI152 template.  
The remaining 32 T1 images were then registered to the
template.  Each DTI and b0 image was mapped to the MNI152 map as 
described above using the T1 to MNI composite mapping.  DTI 
were reoriented using the PPD method \cite{alexander2001}.  The 62
aligned DTI were averaged using the log-euclidean framework 
\cite{arsigny2006} to create the DTI atlas.  Similarly, the 62 aligned
b0 images were averaged to create the representative b0 image
for simulated DTI creation.  The derived FA and b0 images are shown in 
Fig. \ref{fig:simulated}.  An FA mask was created by thresholding 
the FA image of the DTI atlas at 0.2.

The 41 aligned male DTI were used to generate 18 ``Control'' 30-direction
diffusion-weighted images
with no simulated noise nor pathology.  Similarly, the 21 aligned female
DTI were used to generate 19 ``Experimental'' diffusion-weighted images.  
DTI were reconstructed using Camino which were then use to create
individual FA images were
created from the DTI images to use in the simulated experiments.
The number of subjects in each
group was chosen based on the median control and experimental
subject numbers from a sampling of FA studies.

>>>>

\begin{abstract}
Recent discussions within the neuroimaging community have highlighted the
problematic presence of selection bias in experimental design.  Although 
initially centering around the selection of voxels during the course of
fMRI studies, we demonstrate how this concept of circularity bias can 
potentially corrupt voxel-based analysis with a particular emphasis on
population studies of fractional anisotropy (FA).  
For such studies, template-based registration plays a critical role
in which a representative template serves as the normalized space for 
group alignment.  Following 
computation of FA, each subject's FA map is registered to the FA template prior to 
performing statistical comparisons between different groups within a population.
We illustrate how this analysis strategy can induce selection
bias which varies in strength with the similarity metric used in
registration.  Specifically, we analytically demonstrate that the popular sum of squared
difference (SSD) intensity metric explicitly maximizes effect size rather than
anatomical alignment---an experimental design flaw referred to as `circularity bias.'  
Through the use of both simulated diffusion data and real data involving
traumatic brain injury survivors, we explore the magnitude of this bias
and how it varies with similarity metric under the hypothesis that while
SSD-related metrics, such as Demons, will manifest similar effects, 
other metrics which 
are not formulated based on absolute intensity differences 
(e.g. mutual information (MI) and cross correlation (CC)) will produce
less of an effect. 
Consequently, given the variability in voxel-based analysis outcomes 
with similarity metric choice,
we caution researchers specifically in the use of SSD 
and SSD-related measures where normalization and statistical 
analysis involve the same image set and, instead, advocate a more 
cautious approach
where normalization of the individual subject images
to the statistical reference space occurs through the 
corresponding (statistically independent) anatomical images to 
an anatomical template.

%One possibility, the Symmetric Group Normalization
%strategy for unbiased template construction from anatomical (e.g.
%T1-weighted) images, provides a valid and unbiased coordinate system for
%statistical inference of statistical differences.  
%
%
%
%The first set of experiments explores the different outcomes 
%
%
%We establish this
%outcome by using simulated data derived from normal controls from the International
%Neuroimaging Data-sharing Initiative as well as a cohort of 16 traumatic
%brain injury survivors compared with 17 control subjects.  
%Alignment
%comparisons with the normalization component of the popular tract-based
%spatial statistics (TBSS) show that our anatomically-based template
%approach produces good alignments despite the absence of any direct
%FA-to-FA registration.  
%
%
%In conclusion, we recommend against circular analysis strategies that 
%select image similarity metrics that are tied directly to the statistical 
%functions used to assess population effects. 
\end{abstract}

@@

However, suppose the SSD metric is used to find a set of $M$ transformations that map a population of $M$ FA images, $\{I_1, I_2, \ldots, I_M\}$, to a representative FA template, $J$, for statistical testing between groups. This results in the transformations,
$\boldsymbol{\mathcal{T}} = \{\mathcal{T}_1, \mathcal{T}_2,\ldots,\mathcal{T}_M \}$, which minimize the SSD metric over the population, i.e.
\begin{align}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{m=1}^M\sum_{n=1}^N \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2.
\end{align}
Switching the order of the summations (since the transformations are mutually independent) and recognizing that a principal criterion for the selection of $J$ is such that it be a good approximation of the mean of the aligned images, it becomes apparent that the transformation solution
\begin{align}\label{eq:variance}
  \boldsymbol{\mathcal{T}} = \underset{\mathcal{T}_1, \ldots, \mathcal{T}_M}{\operatorname{argmin}}
    \sum_{n=1}^N \underbrace{\sum_{m=1}^M \left( I^n_m( \mathcal{T}_m(\mathbf{x}) ) - J \right)^2}_{
    \propto\mathrm{\,voxelwise\,variance}}
\end{align}
is that which minimizes the voxelwise group variance.  Reducing the group variance will also directly affect the value of a standard population statistic, the Student's $t$-test.  Based on this derivation, one would also expect the popular Demons metric to be susceptible to this erroneous increase in statistical power due to identical objective functions although a more aggressive displacement/gradient at each iteration will relatively exacerbate this effect for a given number of iterations.  
Other standard similarity metrics will also exhibit this
trend to some degree as they minimize some local measure of intensity-based distance between the subject and the template although the effect will be minimized relative to the similarity to Eqn. (\ref{eq:variance}).

@@

Although various processing choices mitigate the effects of such bias,
such as Gaussian smoothing during and following normalization, 
nonparametric testing \citep{Rorden2007}, 
statistical analysis based on orthogonal projections onto the white matter skeleton 
\cite{Smith2006}, multiple comparisons corrections \cite{Nichols2003}, and the use of more robust similarity metrics, 
such choices are often made ad hoc and/or post hoc and, to our 
knowledge, not with the realization of the potential
for circularity bias described by Eqn. (\ref{eq:variance}).

Instead of direct FA-to-FA template mapping for 
alignment to the normalized space, we advocate a statistically 
independent alignment protocol in which the FA images are mapped to 
the common reference space via intra-subject FA-to-T1 mapping 
followed by T1-to-T1 template alignment thus eliminating all direct
FA-to-FA registrations. The key point is that the intensity value differences used 
to assess alignment (i.e. the similarity metric) in FA-to-T1 and T1-to-T1 
normalizations are statistic gally independent of the intensity value differences used for 
the post-alignment numerical analysis.  It should be noted that this discussion
is certainly not limited to studies involving FA.  It should be apparent
that such bias can occur anytime the images used for normalization
are non-independent of the statistical analysis as defined by the similarity
metric and statistical testing.

Two sets of experiments establish that the similarity metric
can significantly affect voxel-based analysis outcomes and that,
based on the previous description of circularity bias,
the relationship in performance variability and similarity metric
is to be expected.  Specifically, we show that whereas SSD and Demons
substantially overestimate statistical significance, mutual
information and cross correlation are less susceptible although the
effect is not
negligible.  Furthermore, we also show that this effects persists
even with the common practice of post-normalization smoothing prior
to mass univariate statistical testing.

Given this general overview, the experimental sets can be briefly
characterized as follows:
\begin{itemize}
  \item {\it Simulated DTI Data}:  Using a simulated DTI generator,
  FA images based on male/female populations were generated in situ
  thus guaranteeing anatomical alignment prior to processing.
  The simulated images in each group were then ``registered'' to the FA image
  of the defining DTI 
  template using four common similarity metrics (CC, Demons, MI, and SSD).  
  ROC curves based on voxel-based analysis for each of these scenarios demonstrate that 
  effect size is maximized over anatomical alignment and that
  this effect size varies with similarity metric.  Additional
  experiments establish that this trend persists even for commonly used smoothing
  kernels and the projection step of TBSS.
  \item {\it Real TBI Data}:  FA data of TBI survivors and controls were 
  normalized to a template using four common similarity metrics
  (CC, Demons, MI, and SSD).  Additional normalization included normalization via
  the subjects' anatomical T1-weighted images as well as the default registration for TBSS.
  Voxelwise statistical testing was also performed with
  and without smoothing.  Even though ground truth is not available for such
  data, the SSD and Demons metrics showed increased statistical power as evidenced
  by a general decrease in voxelwise $p$-values.
\end{itemize}
All major processing steps were performed using the open source Advanced Normalization Tools 
(ANTs)%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
software repository which should facilitate reproducibility for the interested 
reader.

% We then show how the general application
%of a high performance \citep{Klein2009}, nonrigid registration algorithm 
%\citep{Avants2011}, which forms the underlying machinery for an optimal 
%template construction algorithm \citep{Avants2010}, can facilitate alignment 
%between DTI-derived scalar images for cross-population studies.  
%Although this type of alignment procedure is not uncommon 
%(e.g. \cite{Lipton2008}), to our knowledge, this is the 
%first time a connection has been made with the concept of 
%circularity.  
%The rest of the paper establishes the severity of 
%this bias in situations
% although scientific best 
%practices would dictate the reasonable removal of all 
%potential sources of bias (including circularity) regardless of 
%the general degree of severity.

\section{Materials and Methods}
% Provide sufficient detail to allow the work to be reproduced. 
% Methods already published should be indicated by a reference: 
% only relevant modifications should be described.

%\subsection{Template Construction for Anatomically-Based Population Normalization}
%
%\begin{figure*}
%\begin{center}
%\begin{tabular}{c}
%  \includegraphics[width=170mm]{workflow.pdf}
%%  \includegraphics[width=130mm]{workflow.pdf}
%\end{tabular}
%\caption{Possible ANTs-based workflow using SyGN for normalization of images for statistical analysis.  In contrast to standard  FA-to-FA template registration
%(transform represented with the blue arrow), our proposed modification first constructs an ANTs optimal anatomical template as illustrated in the center of the figure.  Each FA image is then rigidly mapped, with distortion correction, to the individual T1-weighted using the white matter mask as shown on the left.  The individual T1-weighted image is then nonrigidly registered to the anatomical template.  Optionally, one can register the T1-weighted template to the MNI152 template which resides in the same space as the FMRIB58 template.  Thus, the composite transform, carefully constructed without any FA-to-FA image registration, can take an individual subject's FA map (or other DTI-derived measure) to the desired standardized coordinate system.    }
%\label{fig:workflow}
%\end{center}
%\end{figure*}
%
%
%
%In general, normalization to a standard coordinate system to
%reduces intersubject variability in population studies.  Various
%approaches exist for determining the normalized space such as selection
%of a pre-existing template based on a single subject, e.g. the Talairach
%atlas \citep{Talairach1988}, or a publicly available averaged group of
%subjects, e.g. the MNI \citep{Collins1994} or ICBM \citep{Mazziotta1995}
%templates.  
%%One challenge with standard templates is that they may inadvertantly bias one's results by enabling better normalization of subjects to which the template is more similar.  This issue is exacerbated when dealing with populations that have high variance (e.g. due to disease) and/or when one's normalization method is low-dimensional (not flexible enough to capture large shape differences). 
%Population-specific templates (or optimal templates) alleviate some of
%these issues by deriving a most representative image from the population
%\citep{Good2001}.  Large deformation registration algorithms also reduce
%this confound by being less sensitive to the deformation distance
%between subject and target.  Some approaches combine both advantages,
%for instance, the diffeomorphic approach of Joshi et al. employs the
%SSD metric and a shape distance to bring the subject group of images
%into alignment \citep{Joshi2004}.  Variants
%include extension to multiple modalities \citep{Lorenzen2006} and small deformations
%\citep{Geng2009}.  These approaches iteratively minimize group difference in ``congealing''
%towards a representative image template \citep{Learned-Miller2006}.
%In contrast, Avants et al. explicitly model the geometric component of the 
%normalized space during optimization.  Coupling the intrinsic symmetry of 
%SyN pairwise registration \citep{Avants2011} and an
%optimized sharpening/averaging of the template appearance, Symmetric Group
%Normalization (SyGN) is a powerful framework for producing optimal population-specific
%templates \citep{Avants2010} with arbitrary similarity metric choice. 
%
%%Given a set of representative images, 
%%$\{I_1, I_2, \ldots, I_M\}$, optimization involves finding the set of paired
%%diffeomorphic transformations, $\left\{\left(\phi^1_1, \phi^1_2\right), 
%%\left(\phi^2_1, \phi^2_2\right), \ldots, \left(\phi^M_1, \phi^M_2\right) \right\}$,
%%the optimal template appearance, $J^*$, and corresponding coordinate system, $\psi(\mathbf{x})$,
%%which minimize the following cost function:
%%\begin{align}
%%  \sum_{m=1}^M \left[
%%    D\left( \psi(\mathbf{x}), \phi^m_1(\mathbf{x}, 1) \right) + 
%%    \Pi \left( I_m\left(\phi^m_2(\mathbf{x}, 0.5) \right), J^*\left(\phi^m_1(\mathbf{x}, 0.5) \right) \right)
%%    \right]
%%\end{align}
%%where $D$ is the diffeomorphic shape distance, 
%%\begin{align}
%%  D\left(\phi(\mathbf{x}, 0), \phi(\mathbf{x}, 1)\right) = \int_{0}^1 \| v(\mathbf{x}, t) \|_L dt, 
%%\end{align}
%%dependent upon the choice of the linear operator, $L$, and
%%$v$ is the diffeomorphism-generating velocity field, 
%%\begin{align}
%%  v\left(\phi(\mathbf{x}, t), t \right) = \frac{d\phi(\mathbf{x}, t)}{dt},\,\,\, \phi(\mathbf{x}, 0) = \mathbf{x}.
%%\end{align}
%%$\Pi$ is the choice of similarity metric, often cross-correlation \citep{Avants2008a}, calculated in the 
%%virtual domain midway between each individual image and the current estimate of the template. 
%%
%%With initial assignment of $\left\{\left(\phi^m_1, \phi^m_2\right)\right\}$ and $\psi(\mathbf{x})$ 
%%to identity, iterative optimization
%%involves estimating the pairwise transformations, estimation of the optimal template appearance, and 
%%updating $\psi(\mathbf{x})$ by averaging the current estimate of $\left\{\phi_1^m\right\}$.  Additional details are given in \cite{Avants2010}.  
%%For the interested reader, a set of scripts has been created and is distributed with the ANTs library for facilitating template creation.
%
%Our proposed workflow applied to anatomically-based population normalization is illustrated in Fig. \ref{fig:workflow}.  Instead of direct FA-to-FA registration, our ANTs-based modification can be described as follows:
%
%\begin{enumerate}
%  \item Build the anatomical template from  the population sample or sample subset using SyGN as described in
%  the previous section.
%  \item If only using a sample subset for template creation, register the remaining
%        anatomical data to the template and save the resulting transformation for each         
%        subject indexed by $i$ (i.e. $\mathcal{T}^i_{template}$).
%  \item Find the optimal transformation from each
%        subject's FA to the anatomical template:
%  \begin{enumerate}
%  \item For each anatomical image, segment the white matter (in ANTs, one can use N4 bias correction \citep{Tustison2010} and Atropos $n$-tissue segmentation \citep{Avants2011a}).
%  \item Rigidly transform each subject's average DWI to the bias-corrected anatomical image ($\mathcal{R}^i_{dwi}$).
%  \item Perform distortion correction on the FA to anatomical mapping 
%        by localized nonrigid registration
%        of rigidly transformed FA map ($=\mathcal{R}^i_{dwi} \mathrm{FA}$) using
%        the white matter mask obtained in step 3(a) ($\mathcal{T}^i_{FA}$).
%  \item For each subject, concatenate the transforms which maps each
%        subject's FA to the template ($\mathcal{T}^i_{total} = \mathcal{T}^i_{template} \circ 
%        \mathcal{T}^i_{FA} \circ \mathcal{R}^i_{dwi} \mathrm{FA}$).
%  \item Warp each subject's FA map to the template using $\mathcal{T}^i_{total}$.  
%  Equivalently, one could warp the mean diffusion (MD), axial diffusion (AD) or any 
%  other scalar derivable from DTI to the template.  
%%  One other detail is that we prefer to warp the subject's DTI to 
%%  the template and then, in template space, derive the scalar images of interest.  
%  \end{enumerate}      
%%  \item Perform the remaining portions of the standard TBSS pipeline:
%%    \begin{enumerate}
%%      \item Create the mean FA image from all template-aligned FA images.
%%      \item Produce the white matter skeleton from the mean FA image.
%%      \item Create skeleton distance map and project all FA data onto the skeleton.
%%      \item Run the statistical analysis.
%%    \end{enumerate}
%\end{enumerate}
%%The next section compares this framework to FA-based normalization strategies with respect to both a patient-control study and a study of gender effects in a control population.

@@

\subsubsection{NKI/Rockland Data for Generation of the Simulated DTI Gender Cohort}
Data from the first 14 weeks of prospective data-sharing initiative sponsored by the 1000 Functional Connectomes Project (FCP)%
\footnote{
In support of open science, the 1000 Functional Connectomes Project (FCP---http://fcon\_1000.projects.nitrc.org) 
was initiated on December 11, 2009 by various members of the MRI community \citep{Biswal2010}.  Motivated by the absence of neuroimaging data for research, this initiative seeks to form collaborative partnerships with imaging institutions for sharing well-documented multi-modal image sets accompanied by phenotypic data.  Commitments from institutions such as the NYU Institute for Pediatric Neuroscience and the Nathan Kline Institute (NKI)--Rockland have resulted in prospective data repositories
currently distributed to the public via the Neuroimaging Informatics Tools and Resources Clearinghouse (NITRC---http://www.nitrc.org).
}
were downloaded on January 15, 2011 comprised of 74 subjects (average age in years: 32.4 $\pm$ 17.8).  
Due to various issues (e.g. lack of corresponding T1-weighted image, failed DTI reconstruction, and age matching requirements) only 62 subjects (21 females and 41 males).  Each imaging session for each subject produced a 64-directional diffusion tensor imaging scan (parameters: conventional single-shot spin echo
EPI pulse sequence, TR = 10000 ms, TE = 91 ms, axial acquisition, 
voxel size = $2 \times 2 \times 2$ mm$^3$) and a T1-weighted anatomical scan (parameters:  TR = 2500 ms, b = 1000 s/mm$^2$ for each direction, 58 contiguous slices of 2.0 mm thickness,
TI = 1200 ms, TE = 3.5 ms, flip angle = 8$^\circ$, 192 contiguous slices of 
1.0 mm thickness, FOV = 256 $\times$ 256 mm$^2$, and voxel size = 1 mm$^3$).  Images were anonymized including defacing of the T1-weighted images prior to upload. Diffusion tensor data associated with each of the following data sets were reconstructed from the diffusion weighted sequences using Camino%
\footnote{
http://camino.org.uk/
}---an open source toolkit for diffusion MRI processing and 
analysis \citep{Cook2006} in combination with ANTs-based 
registration for motion correction of the DTI sequence.

@@

\subsection{Simulated DTI Gender Cohort Creation}
\label{sec:sim}

@@

To provide realistic ground truth FA data, we implemented 
the method of Van Hecke et al. \cite{van-hecke2009} which 
we have also made available to the public
\cite{Tustison2011}.%
\footnote{
http://hdl.handle.net/10380/3315
}  
Given a DTI atlas, a b0 image, a scheme file indicating
the gradient directions, and 
a cohort of aligned DTI data (to model intersubject variability), the
algorithm generates
simulated data sets with specified pathology and noise.
For the simulated experiments described in the next section we
created a T1-weighted template from 30 randomly chosen T1-weighted images of
the NKI/Rockland data set described above.  
The resulting
template was then registered to the MNI152 template.  
The remaining 32 T1-weighted images were also registered to the
template resulting in a composite transform for each subject describing
the transform from the individual subject's T1-weighted image domain to 
the MNI152 template.
